---
title: "Trial mortality predictions - NLST"
author: "Lukas Owens"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning=FALSE, message=FALSE)
library(tidyverse)
source('core.R')
```

### Inputs and staging

Values taken from published trial: 

https://www.nejm.org/doi/full/10.1056/nejmoa1102873
  
https://pubmed.ncbi.nlm.nih.gov/31995683/

```{r}

# Read total cases and death count by arm extracted from trial
trial_counts <- read_csv('../data/trial-data-nlst.csv')

# Subjects were followed for 7 year, 5 months (midpoint of recruitment to end of followup)
trial_length <- (5+7*12)/12

# Person-years in each arm, based on total cases and cases per PY (in abstract)
py_ctrl   <- 941 / 572 * 100000 
py_screen <- 1060 / 645 * 100000 

# Define early vs late stage disease
earlylate <- tribble(
  ~stage, ~late,
  'IA',    'E',  
  'IB',    'E',  
  'IIA',   'E',
  'IIB',   'E',    
  'IIIA',  'A',
  'IIIB',  'A',    
  'IV',    'A'     
)

# Total deaths in control arm
total_deaths_ctrl <- 309

# Actual observed mortality reduction (and upper/lower CI bounds)
actual <- 0.20
lb <- 0.07
ub <- 0.27
```


Group and summarize by early/advanced stage definition
```{r}
trial_counts_ea <- trial_counts %>% 
  left_join(earlylate) %>%
  group_by(arm, late) %>%
  summarise(cases = sum(cases)) %>%
  mutate(case_dist = cases / sum(cases))
trial_counts_ea
```


### Model calculations
Calculate incidence rate ($\lambda_d$), rate of late-stage cancer ($p_1$) and reduction in late-stage ($\alpha$)
```{r}
# Incidence from paper abstract
lambda_d <- 572 / 100000 
lambda_d

adv_cases_ctrl   <- trial_counts_ea %>% filter(arm=='control', late=='A') %>% pull(cases)
adv_cases_screen <- trial_counts_ea %>% filter(arm=='screen', late=='A') %>% pull(cases)
alpha <- 1 - (adv_cases_screen / py_screen) / (adv_cases_ctrl / py_ctrl)
alpha

p_1 <- trial_counts_ea %>% filter(arm=='control', late=='A') %>% pull(case_dist)
p_1
```


Calculate overall mortality hazard implied by case fatality rate 
```{r}
fatality_rate <- 309 / 572 # (Deaths/100k) / (Cases/100k) in control arm
case_hazard <- lambda_d

# Formula for calculating fatality rate from above inputs and a given mortality hazard
fatality_rate_fn <- function(mort_hazard) {
  (1-(mort_hazard*exp(-case_hazard*trial_length) - case_hazard*exp(-mort_hazard*trial_length)) / 
     (mort_hazard - case_hazard)) / (1-exp(-case_hazard*trial_length))
}

# Determine mortality rate that achieves observed fatality rate 
f <- function(x) abs(fatality_rate_fn(x) - fatality_rate)
mort_hazard <- optimize(f, lower=0, upper=5)$minimum
mort_hazard
```
### Calculate calibrated SEER mortality hazards
Import SEER data
```{r}
# Import and clean raw data
seer <- read_csv('../data/seer-mortality-nlst.csv', 
                 skip=1,
                 col_names = c('page', 'stage', 'interval', 'n', 'median', 'surv', 'se', 'lb', 'ub')) %>%
  mutate(surv = surv/100) %>%
  left_join(earlylate) %>%
  select(stage, late, n, surv)
seer

# Group by early / late
seer_grouped <- seer %>%
  filter(!is.na(late)) %>%
  group_by(late) %>%
  summarise(surv = weighted.mean(surv, n), n=sum(n))
  
# Calculate mortality hazards
seer_hazards <- seer_grouped %>% mutate(hazard_raw = -log(surv) / trial_length )
seer_hazards

```

Calculate stage-specific mortality hazards ($\lambda_0$ and $\lambda_1$) calibrated to SEER data 

```{r}
# Function for calculating mortality hazard given an calibration factor
hazard_adj_fn <- function(adj_factor) {
  adj_surv <- seer_hazards %>% 
    mutate(hazard_adj = hazard_raw * adj_factor,
                          surv_adj = exp(-hazard_adj*trial_length)) %>%
    summarise(surv_adj = weighted.mean(surv_adj, n)) %>%
    pull()
  adj_hazard <- -log(adj_surv) / trial_length
  return(adj_hazard)
}

f <- function(x) abs(hazard_adj_fn(x) - mort_hazard)
adj_factor <- optimize(f, lower=0, upper=5)$minimum
adj_factor

lambda_0 <- adj_factor * seer_hazards %>% filter(late=='E') %>% pull(hazard_raw)
lambda_0
lambda_1 <- adj_factor * seer_hazards %>% filter(late=='A') %>% pull(hazard_raw)
lambda_1

```



### Final calculations

```{r}
r_0 <- calc_r(trial_length, lambda_d, lambda_0)
r_1 <- calc_r(trial_length, lambda_d, lambda_1)
rho <- calc_rho(trial_length, r_0, r_1, p_1)
mrel <- rho * alpha

output <- tribble(
  ~trial, ~lambda_d, ~lambda_0, ~lambda_1, ~p_1, ~rho, ~alpha, ~mrel, ~actual, ~lb, ~ub,
  'nlst', lambda_d, lambda_0, lambda_1, p_1, rho, alpha, mrel, actual, lb, ub
)
output
write_csv(output, '../output/trial-calculations-nlst.csv')
```


